The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-05-22 ms.Rmd
Re-organising the structure of the main section

<!-- The Problem -->

@@

This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

<!-- The Solution: ZOOOOOON!  -->
# The `zoon` R package

Versions 0.4-22 and later of the `zoon` R package can be installed directly from CRAN.

### Constructing a `zoon` workflow

The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
Each of these steps is carried out by one or more community-contributed software modules.
For each of these steps, the user selects one or more 'modules' and combined them in a call to `workflow`.

Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow.

<!--
the figures for this panel are generated in powerpoint (I know, I know, but it's quick and easy) and saved in the figs folder.
the powerpoint file itself is not version controlled, but can be access here: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0
-->

@@

@@

* *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
* *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
* *Process* - Processes  applied to the occurrence and covariate data. These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
* *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data. These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
* *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined. The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

@@

#### Crossvalidation

@@

Structure of workflow objects:

@@

@@

We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles and extending them.
Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below, can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)

@@

```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE, fig.height = 9, fig.width = 9}
Feng_Papes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = PrintMap)
```

@@

```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
Feng_Papes_Update <- 
  ChangeWorkflow(workflow = Feng_Papes,
                   output = Chain(InteractiveMap, 
                                  ResponseCurve(cov = 1)))
```

```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = 'hide', fig.height = 3.2}
# combine the printed map and first effect plot

@@

>>>>

@@

This paper introduces version 0.4-22 of the `zoon` R package, which can be dowloaded from CRAN and which encodes the majority of this functionality. First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

# The `zoon` R package

The `zoon` R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules. Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis. Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage. The module types and module combination options are described in more detail below.

<!— link to powerpoint doc: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0-->

@@

####

@@

The five module types, corresponding to the five key steps of an SDM anaysis are:

<!— needs work—>

- *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
- *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
- *Process* - Processes  applied to the occurrence and covariate data. These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
- *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data. These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
- *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined. The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

<!— needs work—>

@@

@@

A `zoonWorkflow` objectStructure of workflow objects:

@@

#### Additional features

##### Cross-validation

<!— needs work—>

##### Sharing workflows

`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform [`figshare.com`](). `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.

@@

We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)

@@

```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = &#39;hide&#39;, results = &#39;hide&#39;, message = FALSE, fig.height = 9, fig.width = 9}
Feng_Papes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = PrintMap)
```

@@

```{r fengChangeWorkflow, eval = TRUE, fig.show = &#39;hide&#39;}
Feng_Papes_Update <- 
  ChangeWorkflow(workflow = Feng_Papes,
                   output = Chain(InteractiveMap, 
                                  ResponseCurve(cov = 1)))
```

```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = &#39;hide&#39;, fig.height = 3.2}
# combine the printed map and first effect plot

@@

<!— point to github development site—>
